Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-14 surfaces-write-up/surface-write-up.tex
middle of minor changes todos

\noindent
Let $e_1, \ldots, e_t \in \Sigma,$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ be a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{n} \pdeg_i$
. Furthermore, every element $\sigma \in \Sigma$ can be written in
a canonical form $\lambda + \sum_{i=1}^{t} w_i e_i,$ with the first
coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i$ and $w_1,
\ldots, w_t \in \bz$.

\todo{define bidegrees $a_i$ and $b_i$}
\end{prop}

\begin{proof}
\todo{define $\pdeg$, $\bida$, and $\bidb$ Aaron: $\pdeg$ was defined in the lemma statement. The $a_i$ were just real numbers, not bidegrees. I'll change the letter.}
By assumption, $\sigma \in \Sigma$ as $\sigma = \sum_{i=1}^{t} r_i e_i$ with $r_i \in \br$. Let $\{r\} := r - \lfloor r \rfloor$ denote the fractional part of $r$. Let $\lambda = \sum_{i=1}^{t}\{r_i\}$. Hence, we can write $\sigma = \lambda + \sum_{i=1}^{t}\lfloor r_i \rfloor e_i.$ Hence, $\sigma$ lies in the $\bz_{\geq 0}$ span of $\lambda, e_1,\ldots, e_t$, which all have first coordinate less than $\sum_{i=1}^{t}r_i$. Therefore, $\Sigma$ is generated by elements whose first coordinate is less than $\sum_{i=1}^{t}r_i$.
\end{proof}

\begin{lem}
\label{lem:composite-map}
Let $D = \sum_{i=1}^{n}\alpha_i D_i$ where $D_i = V(f_i)$. Suppose we have a surjection $\phi: \bk[x_1,x_2,\ldots, x_r] \rightarrow R_D$ given by $x_i \mapsto p_i(f_1, \ldots, f_n),$ where $p_i$ is a monomial in $f_1,\ldots, f_n$. Then, define
\begin{align*}
	\Sigma = \langle u^d y_1^{c_1} \cdots y_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i. \rangle 
\end{align*}
In this case, we can factor $\phi$ through maps $\chi, \psi$ defined by
\todo{Make this have mapsto on bottom}
$$\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}y_1^{c_{i1}} \cdots y_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}
\end{tikzcd}$$
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$ is at most the maximum of the minimal degree of generation of $\ker \chi$ and the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
Note that we have an exact sequence
$$\begin{tikzcd}
0 \ar r & \ker \chi \ar r & \ker \phi \ar r & \ker \psi \ar r & 0.
\end{tikzcd}$$
To see this, note that the left map is an inclusion because $\phi = \chi \circ \psi$. Next, the right map, given by $\chi|_{\ker \phi}:\ker \phi \rightarrow \ker \psi$ is a surjection, again because $\phi = \chi \circ \psi$. The composition of the two maps is $0$ by definition of the kernel of $\chi$. Finally, if $a \in \ker \phi$ and $a$ lies in the kernel of the right map, then by definition $\chi(a) = 0$.
\end{proof}

@@

\begin{example}
\label{eg:radical}
As an example for why we need to add in ghost points for $\bp^2$, consider $\frac{-1}{5}V(x_0^2) + \frac{1}{7}V(x_1^2) + \frac{1}{17}V(x_2^2)$. In degree $595 = 5* 7 * 17$, this has dimension $6$, but we cannot possibly use the lattice approach, as all lower degrees only have dimension 1. In fact, we may construct similar examples using any $x_i^m$ for $f_i$ with appropriate coefficients.
\end{example}
\todo{construct class of additional examples}
\todo{The example above needs to be changed because $x_0^2$ can be replaced by $2x_0$}

>>>>

\noindent
Let $e_1, \ldots, e_t \in \Sigma,$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ be a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{n} \pdeg_i$
. Furthermore, every element $\sigma \in \Sigma$ can be written in
a canonical form $\lambda + \sum_{i=1}^{t} w_i e_i,$ with the first
coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i$ and $w_1,
\ldots, w_t \in \bz$.
\end{prop}

\begin{proof}
By assumption, $\sigma \in \Sigma$ as $\sigma = \sum_{i=1}^{t} r_i e_i$ with $r_i \in \br$. Let $\{r\} := r - \lfloor r \rfloor$ denote the fractional part of $r$. Let $\lambda = \sum_{i=1}^{t}\{r_i\}$. Hence, we can write $\sigma = \lambda + \sum_{i=1}^{t}\lfloor r_i \rfloor e_i.$ Hence, $\sigma$ lies in the $\bz_{\geq 0}$ span of $\lambda, e_1,\ldots, e_t$, which all have first coordinate less than $\sum_{i=1}^{t}r_i$. Therefore, $\Sigma$ is generated by elements whose first coordinate is less than $\sum_{i=1}^{t}r_i$.
\end{proof}

\begin{lem}
\label{lem:composite-map}
Let $D = \sum_{i=1}^{n}\alpha_i D_i$ where $D_i = V(f_i)$. Suppose we have a surjection $\phi: \bk[x_1,x_2,\ldots, x_r] \rightarrow R_D$ given by $x_i \mapsto p_i(f_1, \ldots, f_n),$ where $p_i$ is a monomial in $f_1,\ldots, f_n$. Let $\bida_1, \ldots, \bida_n, \bidb_1, \ldots, \bidb_n \in \bz_{\geq 0}.$ Then, define
\begin{align*}
	\Sigma = \langle u^d y_1^{c_1} \cdots y_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i. \rangle 
\end{align*}
In this case, we can factor $\phi$ through maps $\chi, \psi$ defined by
\todo{Make this have mapsto on bottom}
$$\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}y_1^{c_{i1}} \cdots y_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}
\end{tikzcd}$$
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$ is at most the maximum of the minimal degree of generation of $\ker \chi$ and the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
Note that we have an exact sequence
$$\begin{tikzcd}
0 \ar r & \ker \chi \ar r & \ker \phi \ar r & \ker \psi \ar r & 0.
\end{tikzcd}$$
To see this, note that the left map is an inclusion because $\phi = \chi \circ \psi$. Next, the right map, given by $\chi|_{\ker \phi}:\ker \phi \rightarrow \ker \psi$ is a surjection, again because $\phi = \chi \circ \psi$. The composition of the two maps is $0$ by definition of the kernel of $\chi$. Finally, if $a \in \ker \phi$ and $a$ lies in the kernel of the right map, then by definition $\chi(a) = 0$.
\end{proof}

@@

\begin{eg}
\label{eg:hyperplane}

\end{eg}

\begin{example}
\label{eg:radical}
In this example, we justify the necessity of adding ghost points for the bounds obtained in Lemma ~\ref{P-m-generators}, which are also necessary when the divisors are not all contained in a hyperplanes.

 As an example for why we need to add in ghost points for $\bp^2$, consider $\frac{-1}{5}V(x_0^2) + \frac{1}{7}V(x_1^2) + \frac{1}{17}V(x_2^2)$. In degree $595 = 5* 7 * 17$, this has dimension $6$, but we cannot possibly use the lattice approach, as all lower degrees only have dimension 1. In fact, we may construct similar examples using any $x_i^m$ for $f_i$ with appropriate coefficients.
\end{example}
\todo{construct class of additional examples}
\todo{The example above needs to be changed because $x_0^2$ can be replaced by $2x_0$}
